Supermartingales and Termination Analysis of Statistical PCF

### 2021-04-28 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning. 
The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

>>>>

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
Our definition of AST is equivalent to that given in \cite{DBLP:conf/esop/MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning. 
The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:conf/esop/MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

### 2021-04-28 LICS21/intro.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
A recent result \cite{MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

>>>>

Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
A recent result \cite{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent work \cite{BeutnerO21} %(under double-blind review elsewhere), 
the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

### 2021-04-28 probabilisticProgrammingMartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOPW21}. Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

\paragraph{}
This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

>>>>

The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{DBLP:conf/esop/MakOPW21}. Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

\paragraph{}
This definition of almost sure termination is equivalent to that given in \citep{DBLP:conf/esop/MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:conf/esop/MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi
